---
title: "Ch13 - Dates and Times with lubridate"
output:
  html_document:
    toc: true
    toc_float: true
---

Need to load **lubridate** explicitly since not part of core **tidyverse**.

```{r}
library(tidyverse)
library(lubridate)
library(nycflights13)
```

# Creating Dates/Times

Tibbles refer to dates as <date>, times as <time> and datetime as <dttm>. In base R,
datetimes are `POSIXct` and dates are `date`. There is no native time type in base R.
The **hms** package provides a time datatype.

Current datetimes are available via typical functions.

```{r}
today()
now()
```

## From Strings

Getting from strings to dates or datetimes is super common. In Excel,
we might use the DATEVALUE() function.

```{r}
str_date <- "2017-09-23"
```

```{r}
d1 <- as.POSIXct(str_date) # Using base R
d1
class(d1)
d2 <- ymd(str_date)        # Using lubridate
d2
class(d2)
```

```{r}

```

```{r}
str_date2 <- "5/9/2017" # Intent is May 9, 2017
```

```{r}
as.POSIXct(str_date2) # Using base R
mdy(str_date2)        # Using lubridate
```

Yikes! Checking out `help(as.POSIXct)` reveals that it only tries a few basic

```{r}
as.POSIXct(str_date2, format="%m/%d/%Y")
```

```{r}
str_dt1 <- "2017-01-31 20:11:59"
dt1 <- ymd_hms(str_dt1)
dt1
class(dt1)
```

```{r}
str_dt2 <- "05-09-2017 20:11:59"
dt2 <- mdy_hms(str_dt2)
dt2
class(dt2)
```

```{r}
str_dt3 <- "05-09-2017 20:11:59"
dt3 <- mdy_hms(str_dt2, tz = "EDT")
dt3
class(dt3)
```

## From Individual Components

```{r}
flights %>%
  select(year, month, day, hour, minute)
```

To create dates or datetimes, use `make_date` or `make_datetime`.

```{r}
flights %>%
  select(year, month, day, hour, minute) %>%
  mutate(
    departure = make_datetime(year, month, day, hour, minute)
  )
```

Let's do this with the all the time fields. Since time is in 4 digit integer
format, we'll do some modulus arithmetic to break out the hour and minute.
We'll create a little helper function that can be used in the **dplyr**
`mutate` clause. Notice also the checks for NA and the use of the
`ends_with` function in the `select` clause.

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```

## From other types

```{r}
as_datetime(today())
as_date(now())
```

In Unix systems, the base date or "Unix Epoch" is 1970-01-01.

```{r}
as_datetime(60 * 60 * 10)
as_date(365 * 10 + 2) 
```


### Exercises

1.  What happens if you parse a string that contains invalid dates?

    ```{r, eval = FALSE}
    ymd(c("2010-10-10", "bananas"))
    ```

1.  What does the `tzone` argument to `today()` do? Why is it important?

1.  Use the appropriate lubridate function to parse each of the following dates:

    ```{r}
    d1 <- "January 1, 2010"
    d2 <- "2015-Mar-07"
    d3 <- "06-Jun-2017"
    d4 <- c("August 19 (2015)", "July 1 (2015)")
    d5 <- "12/30/14" # Dec 30, 2014
    ```

## Date-time components

Just like Excel, we can break datetimes up to get pieces such as month or hour.


```{r}
datetime <- ymd_hms("2016-07-08 12:34:56")

year(datetime)
month(datetime)
mday(datetime)

yday(datetime)
wday(datetime)
```

For `month()` and `wday()` you can set `label = TRUE` to return the abbreviated name of the month or day of the week. Set `abbr = FALSE` to return the full name.

```{r}
month(datetime, label = TRUE)
wday(datetime, label = TRUE, abbr = FALSE)
```

```{r}
as_date(datetime)
```

Let's count flights by date and plot.

```{r}
flights_dt %>%
  group_by(flight_date = as_date(dep_time)) %>%
  summarize(
    num_flights = n()
  ) %>%
  ggplot() + geom_line(aes(x=flight_date, y=num_flights))
```

Day of week patterns give the cyclical pattern. Let's just look at Mondays.

```{r}
flights_dt %>%
  filter(wday(dep_time, label = TRUE) == "Mon") %>%
  group_by(flight_date = as_date(dep_time)) %>%
  summarize(
    num_flights = n()
  ) %>%
  ggplot() + geom_line(aes(x=flight_date, y=num_flights))
```

### Rounding

Convenient floor and ceiling functions exist to enable rounding to nearest
unit of time such as a day or week or hour.

```{r}
flights_dt %>% 
  count(week = floor_date(dep_time, "week")) %>% 
  ggplot(aes(week, n)) +
    geom_line()
```

### Setting Components


```{r}
(datetime <- ymd_hms("2016-07-08 12:34:56"))

year(datetime) <- 2020
datetime
month(datetime) <- 01
datetime
hour(datetime) <- hour(datetime) + 1
datetime
```

Alternatively, rather than modifying in place, you can create a new date-time with `update()`. This also allows you to set multiple values at once.

```{r}
update(datetime, year = 2020, month = 2, mday = 2, hour = 2)
```

You can use `update()` to show the distribution of flights across the course of the day for every day of the year: 

```{r}
flights_dt %>% 
  mutate(dep_hour = update(dep_time, yday = 1)) %>% 
  ggplot(aes(dep_hour)) +
    geom_freqpoly(binwidth = 300)
```

Setting larger components of a date to a constant is a powerful technique that allows you to explore patterns in the smaller components.

### Exercises

1.  How does the distribution of flight times within a day change over the 
    course of the year?
    
1.  Compare `dep_time`, `sched_dep_time` and `dep_delay`. Are they consistent?
    Explain your findings.

1.  Compare `air_time` with the duration between the departure and arrival.
    Explain your findings. (Hint: consider the location of the airport.)
    
1.  How does the average delay time change over the course of a day?
    Should you use `dep_time` or `sched_dep_time`? Why?

1.  On what day of the week should you leave if you want to minimise the
    chance of a delay?

1.  What makes the distribution of `diamonds$carat` and 
    `flights$sched_dep_time` similar?

1.  Confirm my hypothesis that the early departures of flights in minutes
    20-30 and 50-60 are caused by scheduled flights that leave early. 
    Hint: create a binary variable that tells you whether or not a flight 
    was delayed.

## Time spans

The **lubridate** package provides *durations*, *periods* and *intervals*.

### Durations

In **lubridate**, subtracting two datetimes gives a difftime object. It's
like a timedelta object in **pandas** (Python library).

How long since the Unix epoch?

```{r}
unix_age_days <- today() - ymd("1970-01-01")
unix_age_days
class(unix_age_days)
```

While difftimes are ambiguous in that they might be in days, weeks, seconds,
minutes or hours, **lubridate** provides a duration object that is 
always in seconds.

```{r}
as.duration(unix_age_days)
```

There's a bunch of duration convenience functions.

```{r}
dminutes(1)
dhours(1)
ddays(1)
dweeks(1)
dyears(1)
```

Since durations are in exact number of seconds, you can get some
strange results.

```{r}
one_pm <- ymd_hms("2016-03-12 13:00:00", tz = "America/New_York")

one_pm
one_pm + ddays(1)
```

### Periods

Periods are designed to work more intuitively in "human" terms.

```{r}
one_pm
one_pm + days(1)
```

Like with durations, there are a bunch of period convenience functions.

```{r}
minutes(1)
hours(1)
days(1)
weeks(1)
years(1)
```

These come in handy for adding periods to dates.

```{r}
now() + hours(12)
```

There's a few oddities in the flights_dt data with arrivals before departures.

```{r}
flights_dt %>%
  filter(arr_time < dep_time)
```

Oops. These are overnight flights. Need to add a day to the arrival time
of such flights. Note the clever use (by HW) of a logical multiplier.

```{r}
flights_dt <- flights_dt %>% 
  mutate(
    overnight = arr_time < dep_time,
    arr_time = arr_time + days(overnight * 1),
    sched_arr_time = sched_arr_time + days(overnight * 1)
  )
```

Now no arrivals before departures.

```{r}
flights_dt %>% 
  filter(overnight, arr_time < dep_time) 
```


### Intervals

An interval is a duration with a starting point. It's used to remedy
ambiguous things like:

```{r}
years(1) / days(1)
```

```{r}
next_year <- today() + years(1)
(today() %--% next_year) / ddays(1)
```

To find out how many periods fall into an interval, you need to use integer division:

```{r}
(today() %--% next_year) %/% days(1)
```

## Time Zones

Dragons in here.

```{r}
Sys.timezone()
```


See Ch 13 in r4ds for more details.